---
title: "Final Project EDA"
subtitle: "ALY 6015"
date: "24 February 2025"
author: "Janica St Ville and Jeff Hackmeister"
format: docx
editor: visual
---

\newpage

# Introduction

For this project, we will be working with a dataset we discovered on Kaggle – The “**Polycystic Ovary Syndrome PCOS”.** This dataset speaks about women with Polycystic Ovary Syndrome and is generated according to the Rotterdam criteria. It sparked our interest because it is a prevalent hormonal disorder that can significantly impact women’s health and quality of life. 

The goal is to explore how various physiological indicators — such as BMI, hormone levels, and menstrual regularity — relate to the presence of PCOS. I hope to identify patterns that may help in early detection or risk assessment. 

To begin the analysis, we will first load in the data from Kaggle and explore the structure of the dataset.

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(patchwork)
# read in CSV from https://www.kaggle.com/datasets/lucass0s0/polycystic-ovary-syndrome-pcos
data <- read.csv("pcos_rotterdam_balanceado.csv")

str(data)
```

The data set contains 3,000 records, 6 variables which are variables of interest including: 

Age – Age of the individual 

BMI – Body Mass Index 

Menstrual_Irregularity – Indicator of menstrual Irregularities 

Testosterone_Level – measured Testosterone level 

Antral_Follicle_count – Count of antral follicles 

PCOS_Diagnosis- PCOS diagnosis (likely 0- NO. 1= yes).

## Data Cleaning

First , we will check for missing values.

```{r}
colSums(is.na(data))
```

To make further analysis easier, we will change the Menstrual_Irregularity and PCOS_Diagnosis variables to a factors, as well as shorten the longer variable names.

```{r}
# Convert to factor with meaningful labels
data$PCOS_Diagnosis <- factor(data$PCOS_Diagnosis, levels = c(0, 1), labels = c("No", "Yes"))

data$Menstrual_Irregularity <- factor(data$Menstrual_Irregularity, levels = c(0, 1), labels = c("No", "Yes"))

data <- data %>%
  rename(T_Level = Testosterone_Level.ng.dL.,
         Men_Irrg = Menstrual_Irregularity,
         AC_Count = Antral_Follicle_Count,
         PCOS_diag = PCOS_Diagnosis)

head(data)
```

To better compare outcomes further in the analysis, we'll split the dataset by the diagnosis variable.

```{r}
y_diag <- data %>%
  filter(data$PCOS_diag == "Yes")
n_diag <- data %>%
  filter(data$PCOS_diag == "No")
```

## Variable Distribution 

We will be investigating the relationship between the variables and a PCOS diagnosis, it is important to understand the distributions of each variable. W

```{r}
#| message: false
#| warning: false
unique_ages <- seq(min(data$Age) - 0.5, max(data$Age) + 0.5, by = 1)
age_plot <- ggplot(data, aes(x=Age)) + 
  geom_histogram(breaks = unique_ages,
                 fill = "lightblue",
                 color = "black") +
  theme_minimal()

bmi_plot <- ggplot(data, aes(x = BMI)) +
  geom_histogram(fill = "lightblue",
                 color = "black") +
  theme_minimal()

mi_plot <- ggplot(data, aes(x=Men_Irrg)) +
  geom_bar(fill = "lightblue")

t_plot <- ggplot(data, aes(x = T_Level)) +
  geom_histogram(fill = "lightblue",
                 color = "black") +
  theme_minimal()

ac_plot <- ggplot(data, aes(x = AC_Count)) +
  geom_histogram(fill = "lightblue",
                 color = "black") +
  theme_minimal()

pcos_plot <- ggplot(data, aes(x=PCOS_diag)) +
  geom_bar(fill = "lightblue")

combined_plot <- (age_plot + bmi_plot) / (ac_plot + t_plot) / (mi_plot + pcos_plot) + 
  plot_layout(guides = "collect")

print(combined_plot)
```

From this we can see a fairly even distribution of ages between 18 and 44, with a slight skew towards the younger ages. The BMI plot produces a mostly normal distribution curve, albeit with a slightly longer tail to the right. The AC count is very left skewed, peaking around 5 and a large drop after 11. Testosterone levels also produced a left skewed distribution with a spike around 50 and a long tale extending well past 100. In the final two plots, you can see the split between the two factor variable.

To take a closer look at the two left skewed variable, we'll look at their distributions by diagnosis as well.

```{r}
#| message: false
#| warning: false
t_plot_y <- ggplot(y_diag, aes(x = T_Level)) +
  geom_histogram(fill = "lightblue",
                 color = "black") +
  theme_minimal()

ac_plot_y <- ggplot(y_diag, aes(x = AC_Count)) +
  geom_histogram(fill = "lightblue",
                 color = "black") +
  theme_minimal()

t_plot_n <- ggplot(n_diag, aes(x = T_Level)) +
  geom_histogram(fill = "lightblue",
                 color = "black") +
  theme_minimal()

ac_plot_n <- ggplot(n_diag, aes(x = AC_Count)) +
  geom_histogram(fill = "lightblue",
                 color = "black") +
  theme_minimal()

(ac_plot_n + t_plot_n) / (ac_plot_y + t_plot_y) + 
  plot_layout(guides = "collect")
```

\newpage

# References

Kottarathil, P. (2020, July 11). Polycystic ovary syndrome (PCOS). Kaggle. https://www.kaggle.com/datasets/prasoonkottarathil/polycystic-ovary-syndrome-pcos
